---
title       : dplyr
subtitle    : 
author      : Almost everything from the tutorial from Kevin Markham
job         : And Hadley Wikham's dplyr resources without which nothing exists
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      # 
widgets     : [mathjax]            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
---

## What is dplyr?

* Tools for data exploration and transformation
* Intuitive to write and easy to read
* Super-fast on data frames

---
## Installing

```{r eval=FALSE}
if (packageVersion("devtools") < 1.6) {
  install.packages("devtools")
}
devtools::install_github("hadley/lazyeval")
devtools::install_github("hadley/dplyr")
install.packages(c("hflights", "Lahman"))
```

---

##  Load our packages

```{r}
suppressMessages(library(dplyr))
library(hflights)
```

hflights: Dataset on commercial domestic flights that departed Houston (IAH and HOU) in 2011.

---
## Our data

```{r}
# explore data
data(hflights)
head(hflights,3)
```

---

## tbl_df

Wraps a data frame that prints nicely

```{r}
flights <- tbl_df(hflights)
head(flights,3) # Can also use print(flights,3) instead
```

---
## Reverting back to regular dataframe
```{r}
# convert to a normal data frame to see all of the columns
head(data.frame(flights),3)
```

---
## Basic single table (df) verbs

1. `filter`: for subsetting variables
2. `select`: for subsetting rows
3. `arrange`: for re-ordering rows
4. `mutate`: for adding new columns
5. `summarise` or `summarize`: for reducing each group to a smaller number of summary statistics 

---

## filter: Keep rows matching criteria

```{r}
# base R: flights[flights$Month==1 & flights$DayofMonth==1, ]
filter(flights, Month==1 & DayofMonth==1)
# same as filter(flights, Month==1, DayofMonth==1)
```

---
## filter again

```{r}
head(filter(flights, UniqueCarrier=="AA" | UniqueCarrier=="UA"),2) 
tail(filter(flights, UniqueCarrier=="AA" | UniqueCarrier=="UA"),2)
```

---
## filter again

```{r}
filter(flights, UniqueCarrier %in% c("AA", "UA"))
```

---
## select: Pick columns by name

```{r}
# base R approach to select DepTime, ArrTime, and FlightNum columns
head(flights[, c("DepTime", "ArrTime", "FlightNum")])
```

---
## select again

```{r}
# dplyr approach
print(select(flights, DepTime, ArrTime, FlightNum),n=6)
```

---
## select again

```{r}
# use colon to select multiple contiguous columns, and use `contains` to match columns by name
# note: `starts_with`, `ends_with`, and `matches` (for regular expressions) can also be used to match columns by name
head(select(flights, Year:DayofMonth, contains("Taxi"), contains("Delay")))
```

---
## Chaining over Nesting?

```{r}
# nesting method to select UniqueCarrier and DepDelay columns and filter for delays over 60 minutes
head(filter(select(flights, UniqueCarrier, DepDelay), DepDelay > 60))
```

---
## Chaining over Nesting

```{r}
flights %>%
    select(UniqueCarrier, DepDelay) %>%
    filter(DepDelay > 60) %>%
    head()
```

---
## Chaining

* Chaining increases readability significantly when there are many commands
* Operator is automatically imported from the [magrittr](https://github.com/smbache/magrittr) package
* Can be used to replace nesting in R commands outside of dplyr

---
## Chaining

```{r}
# create two vectors and calculate Euclidian distance between them
x1 <- 1:5; x2 <- 2:6

# Usual 
sqrt(sum((x1-x2)^2))

# chaining method
(x1-x2)^2 %>% sum() %>% sqrt()
```

---
## arrange: Reorder rows

```{r}
# base R approach to select UniqueCarrier and DepDelay columns and sort by DepDelay
head(flights[order(flights$DepDelay), c("UniqueCarrier", "DepDelay")])
```

---
## arrange

```{r}
# dplyr approach
flights %>%
    select(UniqueCarrier, DepDelay) %>%
    arrange(DepDelay) %>% # arrange(desc(DepDelay)) for descending order
    head()
```

---
## mutate: create new variables that are functions of existing variables

```{r}
# base R approach to create a new variable Speed (in mph)
flights$Speed <- flights$Distance / flights$AirTime*60
head(flights[, c("Distance", "AirTime", "Speed")])
```

---
## mutate

```{r}
# dplyr approach (prints the new variable but does not store it)
flights %>%
    select(Distance, AirTime) %>%
    mutate(Speed = Distance/AirTime*60) %>%
    head()

# store the new variable: flights <- flights %>% mutate(Speed = Distance/AirTime*60)
```

---
## summarise/summarize: Reduce multiple variables to values

* Primarily useful with data that has been grouped by one or more variables
* `group_by` creates the groups that will be operated on
* `summarise` uses the provided aggregation function to summarise each group

```{r}
# base R approaches to calculate the average arrival delay to each destination
head(aggregate(ArrDelay ~ Dest, flights, mean))
# or head(with(flights, tapply(ArrDelay, Dest, mean, na.rm=TRUE)))
```

---
## summarise

```{r}
# dplyr approach: create a table grouped by Dest, and then summarise each group by taking the mean of ArrDelay
flights %>%
    group_by(Dest) %>%
    summarise(avg_delay = mean(ArrDelay, na.rm=TRUE)) %>%
    head()
```

---
## summarise_each/mutate_each: apply the same summary/mutate function(s) to multiple columns at once

```{r}
# for each carrier, calculate the percentage of flights cancelled or diverted
flights %>%
    group_by(UniqueCarrier) %>%
    summarise_each(funs(mean), Cancelled, Diverted) %>%
    head()
```

---
## summarise_each

```{r}
# for each carrier, calculate the minimum and maximum arrival and departure delays
flights %>%
    group_by(UniqueCarrier) %>%
    summarise_each(funs(min(., na.rm=TRUE), max(., na.rm=TRUE)), matches("Delay")) %>%
    head()
```

---
## mutate_each 
```{r}

flights %>%
    select(matches("Delay")) %>%
    head(3)

flights %>%
    select(matches("Delay")) %>%
    mutate_each(funs(half=./2)) %>%
    head(3)
```

---
## n():  counts the number of rows in a group

```{r}
# for each day of the year, count the total number of flights and sort in descending order
flights %>%
    group_by(Month, DayofMonth) %>%
    summarise(flight_count = n()) %>%
    arrange(desc(flight_count)) %>%
    head()
```

```{r eval=FALSE}
# rewrite more simply with the `tally` function
flights %>%
    group_by(Month, DayofMonth) %>%
    tally(sort = TRUE)
```

---
## n_distinct(vector): counts the number of unique items in that vector

```{r}
# for each destination, count the total number of flights and the number of distinct planes that flew there
flights %>%
    group_by(Dest) %>%
    summarise(flight_count = n(), plane_count = n_distinct(TailNum)) %>%
    head()
```

---
## Grouping without summarising

```{r}
# for each destination, show the number of cancelled and not cancelled flights
flights %>%
    group_by(Dest) %>%
    select(Cancelled) %>%
    table() %>%
    head()
```

---
## Window Functions

* Aggregation function (like `mean`) takes n inputs and returns 1 value
* [Window function](http://cran.r-project.org/web/packages/dplyr/vignettes/window-functions.html) takes n inputs and returns n values
* Includes ranking and ordering functions (like `min_rank`), offset functions (`lead` and `lag`), and cumulative aggregates (like `cummean`).

---
## Try this

```{r}
# for each carrier, calculate which two days of the year they had their longest departure delays
# note: smallest (not largest) value is ranked as 1, so you have to use `desc` to rank by largest value
flights %>%
    group_by(UniqueCarrier) %>%
    select(Month, DayofMonth, DepDelay) %>%
    filter(min_rank(desc(DepDelay)) <= 2) %>%
    arrange(UniqueCarrier, desc(DepDelay)) %>%
    head()
```

---
## Other things to play with

```{r eval=FALSE}
# for each carrier, calculate which two days of the year they had their longest departure delays --- rewrite previous with the `top_n` function
flights %>%
    group_by(UniqueCarrier) %>%
    select(Month, DayofMonth, DepDelay) %>%
    top_n(2) %>%
    arrange(UniqueCarrier, desc(DepDelay))

# for each month, calculate the number of flights and the change from the previous month
flights %>%
    group_by(Month) %>%
    summarise(flight_count = n()) %>%
    mutate(change = flight_count - lag(flight_count))

# rewrite previous with the `tally` function
flights %>%
    group_by(Month) %>%
    tally() %>%
    mutate(change = n - lag(n))
```

---
## Other Useful Convenience Functions

```{r eval=FALSE}
# randomly sample a fixed number of rows, without replacement
flights %>% sample_n(5)

# randomly sample a fraction of rows, with replacement
flights %>% sample_frac(0.25, replace=TRUE)

# base R approach to view the structure of an object
str(flights)

# dplyr approach: better formatting, and adapts to your screen width
glimpse(flights)
```

---
## do : for doing arbitrary operations

```{r}
mtcars %>% group_by(cyl) %>% do(head(.,2))
```

```{r eval=FALSE}
# Another example
models <- mtcars %>% group_by(cyl) %>% do(lm = lm(mpg ~ wt, data = .))
models %>% summarise(rsq = summary(lm)$r.squared)
```

---
## Binary verbs

4 joins from SQL

* `inner_join(x, y)`: matching x + y
* `left_join(x, y)` : all x + matching y
* `semi_join(x, y)` : all x with match in y
* `anti_join(x, y)` : all x without match in y

---
## Resources

* [Kevin Markham's tutorial](http://www.dataschool.io/dplyr-tutorial-for-faster-data-manipulation-in-r/)
* [Official dplyr reference manual and vignettes on CRAN](http://cran.r-project.org/web/packages/dplyr/index.html): vignettes are well-written and cover many aspects of dplyr
* [July 2014 webinar about dplyr (and ggvis) by Hadley Wickham](http://pages.rstudio.net/Webinar-Series-Recording-Essential-Tools-for-R.html) and related [slides/code](https://github.com/rstudio/webinars/tree/master/2014-01): mostly conceptual, with a bit of code
* [dplyr tutorial by Hadley Wickham](https://www.dropbox.com/sh/i8qnluwmuieicxc/AAAgt9tIKoIm7WZKIyK25lh6a) at the [useR! 2014 conference](http://user2014.stat.ucla.edu/): excellent, in-depth tutorial with lots of example code (Dropbox link includes slides, code files, and data files)
* [dplyr GitHub repo](https://github.com/hadley/dplyr) and [list of releases](https://github.com/hadley/dplyr/releases)
